---
title: "Codebook"
author: "mFriz"
date: "Tuesday, February 10, 2015"
output: html_document
---

# Context

__Human Activity Recognition Using Smartphones Data Set__ 

http://archive.ics.uci.edu/ml/datasets/Human+Activity+Recognition+Using+Smartphones

The experiments have been carried out with a group of 30 volunteers within an age bracket of 19-48 years. Each person performed six activities (WALKING, WALKING-UPSTAIRS, WALKING-DOWNSTAIRS, SITTING, STANDING, LAYING) wearing a smartphone (Samsung Galaxy S II) on the waist. Using its embedded accelerometer and gyroscope, 3-axial linear acceleration and 3-axial angular velocity were captured at a constant rate of 50Hz. The experiments have been video-recorded to label the data manually. The obtained dataset has been randomly partitioned into two sets, where 70% of the volunteers was selected for generating the training data and 30% the test data.

The sensor signals (accelerometer and gyroscope) were pre-processed by applying noise filters and then sampled in fixed-width sliding windows of 2.56 sec and 50% overlap (128 readings/window). The sensor acceleration signal, which has gravitational and body motion components, was separated using a Butterworth low-pass filter into body acceleration and gravity. The gravitational force is assumed to have only low frequency components, therefore a filter with 0.3 Hz cutoff frequency was used. From each window, a vector of features was obtained by calculating variables from the time and frequency domain. See 'features_info.txt' for more details. 

## Feature Selection
(taken from features_info.txt)

The features selected for this database come from the accelerometer and gyroscope 3-axial raw signals tAcc-XYZ and tGyro-XYZ. These time domain signals (prefix 't' to denote time) were captured at a constant rate of 50 Hz. Then they were filtered using a median filter and a 3rd order low pass Butterworth filter with a corner frequency of 20 Hz to remove noise. Similarly, the acceleration signal was then separated into body and gravity acceleration signals (tBodyAcc-XYZ and tGravityAcc-XYZ) using another low pass Butterworth filter with a corner frequency of 0.3 Hz. 

Subsequently, the body linear acceleration and angular velocity were derived in time to obtain Jerk signals (tBodyAccJerk-XYZ and tBodyGyroJerk-XYZ). Also the magnitude of these three-dimensional signals were calculated using the Euclidean norm (tBodyAccMag, tGravityAccMag, tBodyAccJerkMag, tBodyGyroMag, tBodyGyroJerkMag). 

Finally a Fast Fourier Transform (FFT) was applied to some of these signals producing fBodyAcc-XYZ, fBodyAccJerk-XYZ, fBodyGyro-XYZ, fBodyAccJerkMag, fBodyGyroMag, fBodyGyroJerkMag. (Note the 'f' to indicate frequency domain signals). 

These signals were used to estimate variables of the feature vector for each pattern:  
'-XYZ' is used to denote 3-axial signals in the X, Y and Z directions.

* tBodyAcc-XYZ
* tGravityAcc-XYZ
* tBodyAccJerk-XYZ
* tBodyGyro-XYZ
* tBodyGyroJerk-XYZ
* tBodyAccMag
* tGravityAccMag
* tBodyAccJerkMag
* tBodyGyroMag
* tBodyGyroJerkMag
* fBodyAcc-XYZ
* fBodyAccJerk-XYZ
* fBodyGyro-XYZ
* fBodyAccMag
* fBodyAccJerkMag
* fBodyGyroMag
* fBodyGyroJerkMag

The set of variables that were estimated from these signals are: 

* mean(): Mean value
* std(): Standard deviation
* mad(): Median absolute deviation 
* max(): Largest value in array
* min(): Smallest value in array
* sma(): Signal magnitude area
* energy(): Energy measure. Sum of the squares divided by the number of values. 
* iqr(): Interquartile range 
* entropy(): Signal entropy
* arCoeff(): Autorregresion coefficients with Burg order equal to 4
* correlation(): correlation coefficient between two signals
* maxInds(): index of the frequency component with largest magnitude
* meanFreq(): Weighted average of the frequency components to obtain a mean frequency
* skewness(): skewness of the frequency domain signal 
* kurtosis(): kurtosis of the frequency domain signal 
* bandsEnergy(): Energy of a frequency interval within the 64 bins of the FFT of each window.
* angle(): Angle between to vectors.

Additional vectors obtained by averaging the signals in a signal window sample. These are used on the angle() variable:

* gravityMean
* tBodyAccMean
* tBodyAccJerkMean
* tBodyGyroMean
* tBodyGyroJerkMean


# Getting Data

Courtesy to "Lichman, M. (2013). UCI Machine Learning Repository [http://archive.ics.uci.edu/ml]. Irvine, CA: University of California, School of Information and Computer Science." for providing the data set repository. 

The direct link to the data set is the following:
http://archive.ics.uci.edu/ml/machine-learning-databases/00240/UCI%20HAR%20Dataset.zip

The above .zip file was downloaded and extracted to "UCI-HAR-Dataset". Notice how the empty spaces were replaced with "-" to avoid empty spaces in file path names.  

## File content
(mainly taken from README.txt)

### Main folder
* 'README.txt'

* 'features_info.txt': Shows information about the variables used on the feature vector.

* 'features.txt': List of all features.

* 'activity_labels.txt': Links the class labels with their activity name.

* 'train/X_train.txt': Features of training set.

* 'train/y_train.txt': Train labels (i.e. which activity was actually performed).

* 'test/X_test.txt': Features of testing set.

* 'test/y_test.txt': Test labels (i.e. which activity was actually performed).

### Subfolder
The following files are available for the train and test data. Their descriptions are equivalent. 

* 'train/subject_train.txt': Each row identifies the subject who performed the activity for each window sample. Its range is from 1 to 30. 

* 'train/Inertial Signals/total_acc_x_train.txt': The acceleration signal from the smartphone accelerometer X axis in standard gravity units 'g'. Every row shows a 128 element vector. The same description applies for the 'total_acc_x_train.txt' and 'total_acc_z_train.txt' files for the Y and Z axis. 

* 'train/Inertial Signals/body_acc_x_train.txt': The body acceleration signal obtained by subtracting the gravity from the total acceleration. 

* 'train/Inertial Signals/body_gyro_x_train.txt': The angular velocity vector measured by the gyroscope for each window sample. The units are radians/second. 

### Notes
* Features are normalized and bounded within [-1,1].

* Each feature vector is a row on the text file.

## Merging data
"X_test.txt", "y_test.txt", "subject_test.txt" and the respecting files in the train-folder were merged into a single dataset. 

```{r}
library(data.table)
```
```{r cache=TRUE}
trainData <- read.table(file = "UCI-HAR-Dataset/train/X_train.txt")
trainData <- cbind(trainData, read.table(file = "UCI-HAR-Dataset/train/subject_train.txt", colClasses = "factor"))
trainData <- cbind(trainData, read.table(file = "UCI-HAR-Dataset/train/y_train.txt", colClasses = "factor"))
testData <- read.table(file = "UCI-HAR-Dataset/test/X_test.txt")
testData <- cbind(testData, read.table(file = "UCI-HAR-Dataset/test/subject_test.txt", colClasses = "factor"))
testData <- cbind(testData, read.table(file = "UCI-HAR-Dataset/test/y_test.txt", colClasses = "factor"))                  

dataSet <- rbind(trainData, testData)
```

```{r}
head(colnames(dataSet))
```

## Giving column names
Column names are taken from "features.txt". 

```{r}
features <- read.table(file = "UCI-HAR-Dataset/features.txt", 
                       colClasses = c("character"), 
                       row.names = 1)
colnames(dataSet) <- c(t(features), "subject", "activity")
```
```{r}
head(colnames(dataSet))
```

# Manipulating Data
## Extract only the mean and standard deviation for each measurement. 
Feature names for measurements on the mean contain "-mean()" and feature names for measurements on the standard deviation contain "-std()". Furthermore the columns "subject" and "activity" remain. 
Before we have `r ncol(dataSet)` columns. 

```{r}
columnSelectionVector <- grepl("mean\\(\\)|std\\(\\)|subject|activity", names(dataSet))
dataSetSelection <- dataSet[ , columnSelectionVector]
head(names(dataSetSelection))
```

After this manipulation we are left with `r ncol(dataSetSelection)` columns. 

## Use descriptive activity names to name the activities in the data set.
Activities are the outcomes ("y-values") of the data set. 
* 1 means "WALKING"
* 2 means "WALKING_UPSTAIRS"
* 3 means "WALKING_DOWNSTAIRS"
* 4 means "SITTING"
* 5 means "STANDING"
* 6 means "LAYING"

```{r}
activities <- read.table(file = "UCI-HAR-Dataset/activity_labels.txt", 
        colClasses = c("factor"), col.names = c("activity", "Activity") )

library(plyr)
dataSetSelection <- join(x = dataSetSelection, y = activities, by = "activity", type = "inner")    
dataSetSelection <- dataSetSelection[ , -68]
head(dataSetSelection[ , 68])
```

## Appropriately label the data set with descriptive variable names. 
Names of dataSetSelection so far: 
```{r}
colnames(dataSetSelection[ , c(1,2,3,4,5, 67,68)])

```

The following changes are performed on the column names:
* remove "()"
* replace t with Time
* replace f with Frequency
* replace mean with Mean
* replace std with StdDev (standard deviation)
* capitalize activity and subject


```{r}
colnames(dataSetSelection) <- gsub(pattern = "([()])", replacement = "", x = colnames(dataSetSelection))
colnames(dataSetSelection) <- gsub(pattern = "^t", replacement = "Time-", x = colnames(dataSetSelection))
colnames(dataSetSelection) <- gsub(pattern = "^f", replacement = "Frequency-", x = colnames(dataSetSelection))
colnames(dataSetSelection) <- gsub(pattern = "^activity", replacement = "Activity", x = colnames(dataSetSelection))
colnames(dataSetSelection) <- gsub(pattern = "^subject", replacement = "Subject", x = colnames(dataSetSelection))
colnames(dataSetSelection) <- gsub(pattern = "-mean", replacement = "-Mean", x = colnames(dataSetSelection))
colnames(dataSetSelection) <- gsub(pattern = "-std", replacement = "-StdDev", x = colnames(dataSetSelection))
```

```{r}
head(colnames(dataSetSelection))
```

# Create tidy data set
Create a second, independent tidy data set with the average of each variable for each activity and each subject.
Data.frame structure:
* column 1: Subject
* column 2: Activity
* column 3: -xxx: average of corresponding feature

```{r}
dataSetSelection <- dataSetSelection[ , c(68,67, seq(1,66))] # swap the columns
newTidySet <- ddply(dataSetSelection, c("Subject","Activity"), numcolwise(mean))
print(newTidySet[1:4, 1:4])
write.table(x = newTidySet, file = "TidySet.txt", row.names = FALSE)
```

At this point, the newTidySet contains only average, obtained for each combination of subject and activity. 
